---
title: "Doing Survival Analysis in R"
author: "Mick Cooney"
date: "June 2016"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
    css: styles.css
  pdf_document: default
---

<!--
(Title:) Doing Survival Analysis in R

Author: Mick Cooney

Date: June 2016

Abstract: This Rmarkdown document provides some basic introductory survival analysis in R.

Keywords: survival analysis, intro

-->

```{r knit_opts, include = FALSE}
knitr::opts_chunk$set(
    tidy       = FALSE
   ,cache      = FALSE
   ,fig.width  = 11
   ,fig.height = 8
)

library(data.table)
library(ggplot2)
library(scales)

library(DBI)

library(survival)

library(GGally)

options(width = 80)
```

---

# Introduction

This document illustrates some basic uses of survival analysis for
modelling lapses on a book of life insurance policies.

We assume a basic understanding of survival analysis. For further information on this topic, see [Jon Sedar's blogpost series](http://blog.applied.ai/survival-analysis-part1/).

# Load Data

```{r load_data, echo=TRUE}
dt_protection <- readRDS('dt_protection.rds')
dt_inforce    <- dt_protection[policy_status == 'inforce']

head(dt_protection)

head(dt_inforce)
```

# Create Kaplan-Meier Estimates

First we build a few different KM estimates: for the overall dataset,
for gender and for smoker.

```{r basic_km_overall, echo=TRUE}
lapse_km_overall <- survfit(Surv(durmaxrch, lapsed) ~ 1, data = dt_protection)

ggsurv(lapse_km_overall, plot.cens = FALSE, CI = FALSE) +
    expand_limits(y = 0) +
    ggtitle("Kaplan-Meier Estimate for Life Policies")
```

Now we look at gender

```{r basic_km_gender, echo=TRUE}
lapse_km_gender <- survfit(Surv(durmaxrch, lapsed) ~ gender_life1, data = dt_protection)

ggsurv(lapse_km_gender, plot.cens = FALSE) +
    expand_limits(y = 0) +
    ggtitle("Kaplan-Meier Estimate for Life Policies by Gender")
```

And finally smoker

```{r basic_km_smoker, echo=TRUE}
lapse_km_smoker <- survfit(Surv(durmaxrch, lapsed) ~ smoker_life1, data = dt_protection)

ggsurv(lapse_km_smoker, plot.cens = FALSE) +
    expand_limits(y = 0) +
    ggtitle("Kaplan-Meier Estimate for Life Policies by Smoker")
```


# Build Basic Cox-PH Model

We build some basic Cox-PH models and look at the output. We will also
look at various splits of the data and see if we can induce the
effects of time-varying parameters by splitting the data into groups.

```{r basic_coxph, echo=TRUE}
surv_formula <- formula(Surv(durmaxrch, lapsed) ~ gender_life1 + smoker_life1 +
    mortgage_status + cluster_id + prem_ape)

lapse_coxph <- coxph(surv_formula, data = dt_protection)

summary(lapse_coxph)
```

```{r basic_coxph_plot, echo=TRUE}
dt_baseline <- data.table(gender_life1 = 'F'
                         ,smoker_life1 = 'N'
                         ,mortgage_status = '00TERM'
                         ,cluster_id      = 'n6_c0'
                         ,prem_ape        = mean(dt_protection$prem_ape))

ggsurv(survfit(lapse_coxph, newdata = dt_baseline)
              ,plot.cens = FALSE, CI = FALSE) +
    expand_limits(y = 0) +

    ggtitle("Calculated Hazard Rate for Policy Using Cox-PH Model")
```

`ggsurv` does not seem to work for CoxPH objects like `plot` does in
base R, so instead we can use the output of survfit directly to plot
the survival curves - this might be worth trying to contribute to
GGally at some point.

```{r coxph_curve_plot, echo=TRUE}
lapse_curves <- survfit(lapse_coxph, newdata = dt_inforce)$surv

lapse_curves <- rbind(rep(1, ncol(lapse_curves))
                     ,lapse_curves)

colnames(lapse_curves) <- dt_inforce$policy_id

ggplot(data = melt(lapse_curves[,1:10])) +
    geom_line(aes(x = Var1 - 1, y = value, group = Var2, colour = Var2)) +
    expand_limits(y = 0) +
    xlab("In-force Month") +
    ylab("Cumulative In-force Probability") +
    guides(color = guide_legend(title = 'Policy Id')) +
    ggtitle("Individual Lapse Curves for Sample of Policies")
```



## Look at Strata

One aspect of the Cox-PH model that is also worth exploring is the use
of 'stratified' analysis: this is where we use different baseline
hazard rates for different groupings of the data based on some
categorical data.

How much of an impact does this have on our output?

First we stratify the Cox-PH model by gender

```{r sa_strata_gender, echo=TRUE, results='show'}
lapse_strata_gender_coxph <- coxph(Surv(durmaxrch, lapsed) ~ strata(gender_life1) +
                                       smoker_life1 + mortgage_status + cluster_id +
                                       prem_ape
                                  ,data = dt_protection)

summary(lapse_strata_gender_coxph)

## ggsurv(survfit(lapse_strata_gender_coxph, newdata = dt_baseline)
##               ,plot.cens = FALSE, CI = FALSE) +
##     expand_limits(y = 0) +
##     ggtitle("Calculated Hazard Rate for Policy Using Cox-PH Model")

```

Now we stratify by smoker

```{r sa_strata_smoker, echo=TRUE, results='show'}
lapse_strata_smoker_coxph <- coxph(Surv(durmaxrch, lapsed) ~ gender_life1 +
                                       strata(smoker_life1) + mortgage_status +
                                       cluster_id + prem_ape
                                  ,data = dt_protection)

summary(lapse_strata_smoker_coxph)

## ggsurv(survfit(lapse_strata_gender_coxph, newdata = dt_baseline)
##               ,plot.cens = FALSE, CI = FALSE) +
##     expand_limits(y = 0) +
##     ggtitle("Calculated Hazard Rate for Policy Using Cox-PH Model")

```

Now we stratify by gender and smoker

```{r sa_strata_both, echo=TRUE, results='show'}
lapse_strata_both_coxph <- coxph(Surv(durmaxrch, lapsed) ~ strata(gender_life1) +
                                       strata(smoker_life1) + mortgage_status +
                                       cluster_id + prem_ape
                                ,data = dt_protection)

summary(lapse_strata_both_coxph)

## ggsurv(survfit(lapse_strata_gender_coxph, newdata = dt_baseline)
##               ,plot.cens = FALSE, CI = FALSE) +
##     expand_limits(y = 0) +
##     ggtitle("Calculated Hazard Rate for Policy Using Cox-PH Model")

```



# Time-Dependent Cox-PH by Data Splitting

We also want to split each of the periods of time into blocks of 12
months at a time and then run similar models on each one. The function
`survSplit` does most of the heavy lifting on this.


```{r rundata, echo=TRUE}
dt_split <- survSplit(Surv(durmaxrch, lapsed) ~ .
                     ,data = dt_protection
                     ,cut = 12 * 1:35
                     ,start   = 'start_month'
                     ,episode = 'block_id'
                      )
setDT(dt_split)

dt_split <- dt_split[, .(policy_id
                        ,start_month
                        ,stop_month  = durmaxrch
                        ,lapsecol    = lapsed
                        ,block_id)]

dt_splitdata <- merge(dt_protection
                     ,dt_split
                     ,by = 'policy_id')

dim(dt_splitdata)

head(dt_splitdata)
```

Now that we have built our dataset to split out the different time
blocks, I run the same CoxPH model on each separate block of data:

```{r run_coxph_timeblock}
timedata <- sapply(1:10, function(iterid) {
    block_coxph <- coxph(Surv(start_month, stop_month, lapsecol) ~
                             gender_life1 + smoker_life1 + mortgage_status +
                             cluster_id + prem_ape
                        ,data = dt_splitdata[block_id == iterid])

    return(coef(block_coxph))
})

ggplot(data = melt(timedata)) +
    geom_line(aes(x = Var2, y = value)) +
    facet_wrap(~Var1, scales = 'free') +
    expand_limits(y = 0)
```
