---
title: "FFT LG"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(plotly)
library(ggplot2)
library(tidyverse)
library(mice)
library(tidyr)
library(reshape2)
library(cowplot)


wave_gen=function(amp=1,f=1,t=seq(0,10,length.out = 100),noise_amp=0){
  noise=noise_amp*runif(n=length(t),min=-1,max=1)
  y=(amp*sin(t*2*pi*f))+noise
  return(y)
}
FFT_test=function(amps,fqs,t_vec,noise_amp=0,plot_out=F){
  library(ggplot2)
  waves=matrix(ncol = length(amps),nrow  = length(t_vec))
  sampling=diff(t_vec)[1]
  
  for(n in seq(length(amps))){
    waves[,n]=wave_gen(amp=amps[n],f = fqs[n],t=t_vec,noise_amp)
  }
  
  waves=as.data.frame(waves)
  w_names=""
  
  for(n in seq(length(fqs))){
    w_names[n]=paste0("wave_",n)
  }
  names(waves)=w_names
  
  ggwaves=cbind(waves,t_vec)
  ggwaves$wave_all=rowSums(waves)
  
  p1=ggplot(data=ggwaves,aes(x=t_vec,y=wave_all)) + geom_line()  + ggtitle ("full sample for FFT")
  
  melt_waves = melt(ggwaves, id = "t_vec")
  p2=ggplot(melt_waves %>% filter(variable!="wave_all") , aes(x = t_vec, y = value, group = variable, colour = variable)) +  
    geom_line(size=2) + xlim(0,.1) + ggtitle ("each individual wave")
  
  
  
  N=length(ggwaves$wave_all)
  fft_y=fft(ggwaves$wave_all)
  fft_y=abs(fft_y[1:floor(N/2)])
  fft_y=fft_y/length(t_vec)*2 #amplitude correction
  
  
  nyquist=0.5
  freq=1/sampling*(1:(N/2))/(N/2)*nyquist #sampling adjustment applied
  
  FFT_DF=data.frame(freq,fft_y)
  
  f1=ggplot(FFT_DF,aes(freq,fft_y))+geom_line()  + ggtitle ("Spectrogram")
  f2=ggplot(FFT_DF,aes((1/freq),fft_y))+geom_line()+scale_x_log10()  + ggtitle ("Periodgram")
  
  if(plot_out==T){
      #print(plot_grid(p2,p1,f1,f2,nrow = 2))
    print(p2)
    print(p1)
    print(f1)
    print(f2)
  }
  
  
  return(FFT_DF)
}
FFT_filter=function(wave_in,t_vec,thresh=1){
  sampling=diff(t_vec)[1]
  ggwaves=data.frame(wave_all=wave_in,t_vec=t_vec)
  N=length(ggwaves$wave_all)
  fft_o=fft(ggwaves$wave_all)
  
  #simple peak detection
  peaks_max=max(abs(fft_o))
  peak_locs=which(abs(fft_o)>peaks_max*thresh)
  #peak_locs=c(peak_locs,peak_locs+1,peak_locs-1)
  
  
  ff_filt=rep(0,N)
  ff_filt[peak_locs]=1
  a_filt=fft_o*ff_filt
  a_inv=fft(a_filt,inverse = TRUE)/N
  result=Re(a_inv)
  
  #par(mfrow=c(4,1))
  plot(ggwaves$wave_all,type = "l",main = "Input")
  plot(abs(fft_o),type = "l",main = "FFT")
  plot(abs(a_filt),type = "l",main = "Filtered FFT")
  plot(Re(a_inv),type = "l",main = "Reconstructed")
  #par(mfrow=c(1,1))
  
  
  
  #print(plot_grid(p2,p1,f1,f2,nrow = 2))
  
  
  return(result)
}

FFT_auto=function(Data_in,t_vec_in,amps,fqs,t_vec,noise_amp=0){
  library(ggplot2)
  # waves=matrix(ncol = length(amps),nrow  = length(t_vec))
  # 
  # for(n in seq(length(amps))){
  #   waves[,n]=wave_gen(amp=amps[n],f = fqs[n],t=t_vec,noise_amp)
  # }
  # 
  # waves=as.data.frame(waves)
  # w_names=""
  # 
  # for(n in seq(length(fqs))){
  #   w_names[n]=paste0("wave_",n)
  # }
  # names(waves)=w_names
  # 
  # ggwaves=cbind(waves,t_vec)
  # ggwaves$wave_all=rowSums(waves)
  # 
  # p1=ggplot(data=ggwaves,aes(x=t_vec,y=wave_all)) + geom_line()
  # 
  # melt_waves = melt(ggwaves, id = "t_vec")
  # p2=ggplot(melt_waves %>% filter(variable!="wave_all") , aes(x = t_vec, y = value, group = variable, colour = variable)) +
  #   geom_line()
  # 
  # ggwaves$wave_all=Data_in
  
  sampling=as.numeric(mean(diff(t_vec_in),na.rm = T))
  N=length(Data_in)
  
  fft_y=fft(Data_in)
  fft_y=abs(fft_y[1:floor(N/2)])
  fft_y=fft_y/length(t_vec_in)*2 #amplitude correction
  
  
  nyquist=0.5
  freq=1/sampling*(1:(N/2))/(N/2)*nyquist #sampling adjustment applied
  
  FFT_DF=data.frame(freq,fft_y)
  
  f1=ggplot(FFT_DF,aes(freq,fft_y))+geom_line()
  f2=ggplot(FFT_DF,aes((1/freq),fft_y))+geom_line()+scale_x_log10()
  
  print(plot_grid(f1,f2,nrow = 2))
  
  
  return(FFT_DF)
}



```


###Before processing the data 
The data vector should be peroidic with a known sampling period. Any gaps in the data or jumps in the time series should be addressed with inerpolation or setting the data to 0.
For the Eirgrid data the sampling period is 15 with some missing data set to 0.

##FFT process
* Compute the FFT [fft(df$data)]

* process the FFT with abs(fft[1:floor(N/2)])
    + where N is the length input vector

* construct the frequency vector with (1:(N/2))/(N/2)*nyquist) 
    + where nyquist is the sampling limit and set to 0.5
    
* adjust for the amplitude
    + fft/N*2 #amplitude correction



```{r starting example, echo=F}
t_vec=seq(0,2, by =  .001)


noise=0
amps=c(10,5,3,20,7,15)
fqs=c(20,10,30,40,50,90)

waves1=matrix(ncol = length(amps),nrow  = length(t_vec))

for(n in seq(length(amps))){
  waves1[,n]=wave_gen(amp=amps[n],f = fqs[n],t=t_vec,noise)
}

plot(t_vec,rowSums(waves1),type = "l",xlim = c(0,.2))

fftdf=FFT_test(amps,fqs,t_vec,noise,T)
ggplotly(ggplot(fftdf,aes(freq,fft_y))+geom_line() + ggtitle ("Plotly Spectrogram"))
```
original frequencies were

amps=c(5,10,3,20,7,15)

fqs=c(10,20,30,40,50,90)



##FFT filtering


Can you see the pattern? 

can you see the signal?


```{r filtering  example, echo=FALSE}

t_vec=seq(0,2, by =  .001)


noise=40
amps=c(10,5,3,20,7,15)
fqs=c(20,10,30,40,50,90)

waves=matrix(ncol = length(amps),nrow  = length(t_vec))

for(n in seq(length(amps))){
  waves[,n]=wave_gen(amp=amps[n],f = fqs[n],t=t_vec,noise)
}

wave_in=rowSums(waves)
plot(t_vec,wave_in,type = "l",xlim = c(0,.2),main="Signal with noise added")

fftdf=FFT_test(amps,fqs,t_vec,noise,plot_out = F)
ggplotly(ggplot(fftdf,aes(freq,fft_y))+geom_line()+ ggtitle ("Plotly Spectrogram"))

thresh=(5*sd(fftdf$fft_y))/max(fftdf$fft_y)
#thresh=.9

filt=FFT_filter(wave_in = wave_in,t_vec,thresh)

plot(t_vec,wave_in,type = "l",xlim = c(0,.2),main = "Before Filtering")
plot(t_vec,filt,type="l",xlim = c(0,0.2),main = "After Filtering")

plot(t_vec,filt,type="l",xlim = c(0,0.2),main = "comparison (red is original)")
par(new=T)
lines(t_vec,rowSums(waves1),col="red")

```

#Eirgrid Data

```{r Eirgrid Data}

elec=read.csv("c:/Patrick/DataScience/TimeSeries/system_combined.csv")
elec$Time=as.POSIXct(strptime(elec$Time, "%Y-%m-%d %H:%M:%S"))

summary(elec)
dim(elec)[1]





plot(elec$Time[1:200000],(elec$System.Demand.MW[1:200000]),type = "l")

elec_2014_jan = elec %>% filter(Time >= as.Date("2014-01-01") & Time <= as.Date("2014-02-28") )
p = ggplot(elec_2014_jan, aes(x = Time, y=System.Demand.MW))+  geom_line()
ggplotly(p)

elec_2014 = elec %>% filter(Time >= as.Date("2014-02-21") & Time <= as.Date("2014-03-03") )
p = ggplot(elec_2014, aes(x = Time, y=System.Demand.MW))+ geom_line()
ggplotly(p)



elec$System.Demand.MW=elec$System.Demand.MW-mean(elec$System.Demand.MW)
fft_EirGrid=FFT_auto(Data_in = elec$System.Demand.MW[1:200000], t_vec_in = elec$Time[1:200000])
ggplotly(ggplot(fft_EirGrid,aes(freq,fft_y))+geom_line()+labs(x="cycles per min",y="power/MW") + ggtitle("spectrogram (units unadjusted)"))
ggplotly(ggplot(fft_EirGrid,aes(1/freq,fft_y))+geom_line()+labs(x="min per cycle",y="power/MW") + ggtitle("periodogram (units unadjusted)"))
###########
fft_EirGrid_peri=fft_EirGrid
fft_EirGrid_peri$freq=1/fft_EirGrid_peri$freq/60


p <- plot_ly(fft_EirGrid_peri, x = ~freq, y = ~fft_y,type = "scatter", mode="lines") %>%
  layout(title= "periodogram",xaxis = list(title = "Hours per cycle",type="log"))
p

########
fft_EirGrid_peri=fft_EirGrid
fft_EirGrid_peri$freq=1/fft_EirGrid_peri$freq/(60*24*7)


p <- plot_ly(fft_EirGrid_peri, x = ~freq, y = ~fft_y,type = "scatter", mode="lines") %>%
  layout(title= "periodogram",xaxis = list(title = "weeks per cycle",type="log"))
p

```

