---
title: "Introduction to Time Series"
author: "Mick Cooney"
date: "`r Sys.Date()`"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: pygments
    center: true
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
  pdf_document: default
---

<!--
(Title:) Introduction to Time Series

Author: Mick Cooney

Date: 2 November 2016

Abstract: This is a quick introduction to time series, based largely
on a previous time series workshop I created.

Keywords:

-->

```{r knit_opts, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE
                     ,cache = FALSE
                     ,fig.height =  8
                     ,fig.width  = 11)

library(tidyverse)
library(data.table)
library(dtplyr)

library(ggfortify)
library(plotly)


options(width = 80L)

set.seed(42)
```



---

# Introduction to Time Series

Time series occur in almost any field of study that produces
quantitative data. Whenever quantities are measured over time, those
measurements form a time-series, or more formally, a _discrete-time
stochastic_.


One reasonably famous example of a time-series is count of airline
passengers in the US, as seen below. This is a fairly
simple time-series, with measurements taken on a monthly basis over a
number of years, with each datum consisting of a single number,
i.e. this time-series is _univariate_.


```{r load_air_passengers, echo=TRUE}
data(AirPassengers)

str(AirPassengers)
```

Having quickly looked at the data, we now plot it.

```{r plot_air_passengers, echo=TRUE}
plot(AirPassengers)
```

The above plot uses base R packages - which is limited, so we will
look to run a similar plot but using tools built in ggplot2.

We could build this from scratch, but the package `ggfortify` will
help.



```{r ggfortify_air_passengers, echo=TRUE}
autoplot(AirPassengers) +
    expand_limits(y = 0) +
    xlab("Year") +
    ylab("Count of Passengers ('000s)")

```

## Overview


```{r decompose_timeseries, echo=TRUE}
ap_decompose <- decompose(AirPassengers)

autoplot(ap_decompose)
```

Let's try a multiplicative model.

```{r decompose_timeseries_mult, echo=TRUE}
ap_decompose_mult <- decompose(AirPassengers, type = 'multiplicative')

autoplot(ap_decompose_mult)
```

```{r decompose_timeseries_stl, echo=TRUE}
ap_decompose_stl <- stl(AirPassengers, s.window = 'periodic')

autoplot(ap_decompose_stl)
```


# Autocorrelation (Serial Correlation)

## Air Passengers

```{r first_differences, echo=TRUE}
autoplot(AirPassengers) +
    ggtitle("Plot of the Air Passengers")

autoplot(diff(AirPassengers)) +
    ggtitle("Plot of the Diffs of the Air Passengers")
```

## Air Passenger Correlogram

```{r autocorrelation, echo=TRUE}
ap_decomp_resid <- ap_decompose$random[!is.na(ap_decompose$random)]

ap_acf <- acf(ap_decomp_resid, plot = FALSE)

autoplot(ap_acf) +
    ggtitle("Correlogram of the Air Passenger Residuals")


ap_pacf <- pacf(ap_decomp_resid, plot = FALSE)

autoplot(ap_pacf) +
    ggtitle("Partial Correlogram of the Air Passenger Residuals")

autoplot(pacf(diff(AirPassengers), plot = FALSE)) +
    ggtitle("Partial Correlogram of the Differenced Air Passenger Data")
```


# ARMA and ARIMA Models

```{r generate_innovations, echo=TRUE}
innovations <- rnorm(1000, 0, 1)
```


## MA Models

```{r create_ma_timeseries, echo=TRUE}
ma_1 <- arima.sim(list(ma = 0.8), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ma_1), y = as.numeric(ma_1)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)

ma_2 <- arima.sim(list(ma = c(0.4, 0.4)), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ma_2), y = as.numeric(ma_2)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)


output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ma_1), y = as.numeric(ma_1)), colour = 'red') +
    geom_line(aes(x = seq_along(ma_2), y = as.numeric(ma_2)), colour = 'blue') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)

```

## AR Series

Now that we  have created MA series, we look at what the AR series look like.

```{r create_ar_timeseries, echo=TRUE}
ar_1 <- arima.sim(list(ar = 0.8), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ar_1), y = as.numeric(ar_1)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)


ar_2 <- arima.sim(list(ar = c(0.4, 0.4)), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ar_2), y = as.numeric(ar_2)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)


output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ar_1), y = as.numeric(ar_1)), colour = 'red') +
    geom_line(aes(x = seq_along(ar_2), y = as.numeric(ar_2)), colour = 'blue') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```

## ARMA Models

```{r create_arma_series, echo=TRUE}
arma_1_1 <- arima.sim(list(ar = 0.4, ma = 0.4), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100])) +
    geom_line(aes(x = seq_along(ar_1), y = as.numeric(ar_1)), colour = 'red') +
    geom_line(aes(x = seq_along(ma_1), y = as.numeric(ma_1)), colour = 'blue') +
    geom_line(aes(x = seq_along(arma_1_1), y = as.numeric(arma_1_1)), colour = 'green') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```
