---
title: "Introduction to Time Series"
author: "Mick Cooney"
date: "`r Sys.Date()`"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: pygments
    center: true
---

<!--
(Title:) Introduction to Time Series

Author: Mick Cooney

Date: 2 November 2016

Abstract: This is a quick introduction to time series, based largely
on a previous time series workshop I created.

Keywords:

-->

```{r knit_opts, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE
                     ,cache = FALSE
                     ,fig.height =  6
                     ,fig.width  = 11)

library(tidyverse)
library(data.table)
library(dtplyr)

library(ggfortify)
library(plotly)


options(width = 80L)

set.seed(42)
```

## Air Passengers


```{r ggfortify_air_passengers, echo=FALSE}
autoplot(AirPassengers) +
    expand_limits(y = 0) +
    xlab("Year") +
    ylab("Count of Passengers ('000s)")

```

## Time Series Decomposition


```{r decompose_timeseries, echo=FALSE, warning=FALSE}
ap_decompose <- decompose(AirPassengers)

autoplot(ap_decompose)
```

---

Multiplicative Decomposition

```{r decompose_timeseries_mult, echo=FALSE, warning=FALSE}
ap_decompose_mult <- decompose(AirPassengers, type = 'multiplicative')

autoplot(ap_decompose_mult)
```

---

STL Decomposition

```{r decompose_timeseries_stl, echo=FALSE, warning=FALSE}
ap_decompose_stl <- stl(AirPassengers, s.window = 'periodic')

autoplot(ap_decompose_stl)
```


## Serial Correlation

```{r first_differences, echo=FALSE}
autoplot(diff(AirPassengers)) +
    ggtitle("Plot of the Diffs of the Air Passengers")
```

---

Air Passenger Correlogram

```{r autocorrelation, echo=FALSE}
ap_decomp_resid <- ap_decompose$random[!is.na(ap_decompose$random)]

ap_acf <- acf(ap_decomp_resid, plot = FALSE)

autoplot(ap_acf) +
    ggtitle("Correlogram of the Air Passenger Residuals")
```

---

Partial Correlogram

```{r partial_autocorrelation, echo=FALSE}
ap_pacf <- pacf(ap_decomp_resid, plot = FALSE)

autoplot(ap_pacf) +
    ggtitle("Partial Correlogram of the Air Passenger Residuals")
```

---

Correlogram of First Differences

```{r partial_correlogram, echo=FALSE}
autoplot(pacf(diff(AirPassengers), plot = FALSE)) +
    ggtitle("Partial Correlogram of the Differenced Air Passenger Data")
```

## Modelling Time Series

Time series values as $x_t$.

*Innovations* as $w_t$

$$
\text{Noise} \rightarrow \text{MODEL} \rightarrow \text{signal}
$$


## EWMA Models

Exponentially Weighted Moving Average

$$
x_{t+1} = \lambda w_t + (1 - \lambda) x_{t}
$$

Apply recursively:

$$
x_{t+1} = \lambda w_t + \lambda(1 - \lambda) w_{t-1} + (1 - \lambda)^2  x_{t-2}
$$



## ARMA and ARIMA Models

```{r generate_innovations, echo=TRUE}
innovations <- rnorm(1000, 0, 1)
```

## MA Models

$$
\text{MA(p):} x_t = w_t + \sum^{p}_{i=1} \alpha_i w_{t-i}
$$

---

MA(1) Model

```{r create_ma1_timeseries, echo=FALSE}
ma_1 <- arima.sim(list(ma = 0.8), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ma_1), y = as.numeric(ma_1)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```

---

MA(2) Model

```{r create_ma2_timeseries, echo=FALSE}
ma_2 <- arima.sim(list(ma = c(0.4, 0.4)), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ma_2), y = as.numeric(ma_2)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```

---

Look at models against innovations

```{r ma_plots, echo=FALSE}
output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ma_1), y = as.numeric(ma_1)), colour = 'red') +
    geom_line(aes(x = seq_along(ma_2), y = as.numeric(ma_2)), colour = 'blue') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```

## AR Models

$$
\text{AR(q):} x_t = w_t + \sum^{q}_{i=1} \beta_i x_{t-i}
$$


---

AR(1) Model

```{r create_ar1_timeseries, echo=TRUE}
ar_1 <- arima.sim(list(ar = 0.8), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ar_1), y = as.numeric(ar_1)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```

---

AR(2) Model

```{r create_ar2_timeseries, echo=TRUE}
ar_2 <- arima.sim(list(ar = c(0.4, 0.4)), innov = innovations, n = 100)

output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ar_2), y = as.numeric(ar_2)), colour = 'red') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```


```{r plot_ar_series, echo=FALSE}
output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100]), size = 0.5) +
    geom_line(aes(x = seq_along(ar_1), y = as.numeric(ar_1)), colour = 'red') +
    geom_line(aes(x = seq_along(ar_2), y = as.numeric(ar_2)), colour = 'blue') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```

## ARMA Models

Combine both

$$
\text{ARMA(p,q):} x_{t} = w_t + \sum^{p}_{i=1} \alpha_i w_{t-i} + \sum^{q}_{j=1} \beta_j x_{t-j}


```{r create_arma_series, echo=TRUE}
arma_1_1 <- arima.sim(list(ar = 0.4, ma = 0.4), innov = innovations, n = 100)
```

```{r plot_arma_series, echo=FALSE}
output_plot <- ggplot() +
    geom_line(aes(x = seq_along(innovations[1:100]), y = innovations[1:100])) +
    geom_line(aes(x = seq_along(ar_1), y = as.numeric(ar_1)), colour = 'red') +
    geom_line(aes(x = seq_along(ma_1), y = as.numeric(ma_1)), colour = 'blue') +
    geom_line(aes(x = seq_along(arma_1_1), y = as.numeric(arma_1_1)), colour = 'green') +
    xlab("Time Step") +
    ylab("Value")

ggplotly(output_plot)
```
