---
title: "UK National Lottery Charitable Donations"
author: "Mick Cooney"
date: "28 August 2016"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
  pdf_document: default
---

<!--
(Title:)

Author: Mick Cooney

Date: 2016

Abstract:

Keywords:

-->

```{r knit_opts, include = FALSE}
rm(list = ls())

knitr::opts_chunk$set(
    tidy  = FALSE
   ,cache = FALSE
   ,fig.height =  8
   ,fig.width  = 11
   )

library(data.table)
library(dplyr)
library(dtplyr)
library(ggplot2)

library(readr)
library(feather)
library(readxl)


options(width = 90)

source("custom_functions.R")
```

# Load Data

This data has came in a number of different formats, so the first thing I have done is taken the individual

```{r load_data, echo=TRUE}
data_files <- dir("data/data/", pattern = '\\.csv', full.names = TRUE)

lottery_dt <- rbindlist(lapply(data_files, fread, colClasses = c(COMPANY_NUMBER = 'character')))

setnames(lottery_dt, fix_column_names(names(lottery_dt)))

lottery_dt[, announcement_date    := as.Date(announcement_date,    format = '%d/%m/%Y')]
lottery_dt[, applicant_start_date := as.Date(applicant_start_date, format = '%d/%m/%Y')]
lottery_dt[, decision_date        := as.Date(decision_date,        format = '%d/%m/%Y')]
lottery_dt[, conv_current_award   := as.numeric(gsub("(\\d+) .*", "\\1", current_award))]

glimpse(lottery_dt)


### Read in the summary geospatial data from the XLSX file
summgeo_dt <- read_excel("data/Big Lottery Dataset.xlsx"
                        ,sheet = 'LSOA Shape file'
                        ,col_types = c('text','text','text','text','text','text'
                                      ,'text','numeric','numeric','numeric')
                        )
setDT(summgeo_dt)

setnames(summgeo_dt, fix_column_names(names(summgeo_dt)))

glimpse(summgeo_dt)
```

# Feature Creation

We create a number of derived features from the data.

```{r feature_creation, echo=TRUE}
lottery_dt <- lottery_dt %>%
    mutate(announcement_delay      = as.numeric(announcement_date - decision_date)
          ,lifetime_at_award_days  = as.numeric(decision_date - applicant_start_date)
          ,lifetime_at_award_years = round(as.numeric(decision_date - applicant_start_date) / 365.25, 0)
          )
```

# Univariate Exploration

We start with our standard approach of splitting the features into
groups and then creating univariate plots of the data, the form of
which depends on the variable type.

```{r split_data_types, echo=TRUE}
vars_num <- c("conv_current_award","announcement_delay","lifetime_at_award_days"
             ,"lifetime_at_award_years")
vars_cat <- c("applicant_name","charitable_status","organisation_type"
             ,"programme_name","funder_name","la_area","region","constituency")
vars_dte <- c("applicant_start_date","decision_date","announcement_date")
```

## Numerical Features

First we look at the numerical features, doing a kernel density plot for each of the variables.

```{r plot_numerical_features, echo=TRUE}
for(plot_var in vars_num) {
    cat(paste("Variable: ", plot_var, "\n"))

    count_na     <- sum(is.na(lottery_dt[[plot_var]]))

    var_plot <- ggplot() +
        geom_histogram(aes(x = lottery_dt[[plot_var]]), bins = 50) +
        scale_x_continuous(labels = scales::comma) +
        scale_y_continuous(labels = scales::comma) +
        xlab("Value") +
        ylab("Count") +
        ggtitle(paste0("Plot Variable: ", plot_var
                      ," (", count_na, " NA vals)"
                      )) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

    plot(var_plot)
}
```

## Categorical Features

For categorical variables we do bar plots for each of the levels of the variable

```{r plot_categorical_features, echo=TRUE}
for(plot_var in vars_cat) {
    cat(paste("Variable: ", plot_var, "\n"))

    count_unique <- length(unique(lottery_dt[[plot_var]]))
    count_na     <- sum(is.na(lottery_dt[[plot_var]]))

    var_plot <- ggplot() +
        geom_bar(aes(x = lottery_dt[[plot_var]])) +
        scale_y_continuous(labels = scales::comma) +
        xlab("Value") +
        ylab("Count") +
        ggtitle(paste0("Plot Variable: ", plot_var
                      ," (", count_unique, " unique vals - "
                      ,count_na, " NA vals)"
                      )) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

    plot(var_plot)
}
```

## Date Variables

For date variables we do bar plots for the years in the dates

```{r plot_date_features, echo=TRUE}
for(plot_var in vars_dte) {
    cat(paste("Variable: ", plot_var, "\n"))

    var_plot <- ggplot() +
        geom_bar(aes(x = year(lottery_dt[[plot_var]]))) +
        scale_y_continuous(labels = scales::comma) +
        xlab(paste0("Plot Variable: ", plot_var)) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

    plot(var_plot)
}
```


---
